Errorless irrationality: removing error-driven components from the inverse base-rate effect paradigm

### 2023-01-27 docs/manuscript.tex
transitional commit

The most prominent theories of the inverse base-rate effect involve an an attentional meachanism that drives both learning and responding.
These theories are formal models.
They are: a neural network with exemplar-mediated attention to distinctive input, EXIT \cite{kruschke2001toward}, a three-layer neural network with competitive attentional gating, and a four-layer neural network with an additional rapid attentional shift \cite{paskewitz2020dissecting}.
All these explanations rely on a process that relocates attention in response to prediction errors - they update attentional values according to gradient descent.
Their explanation is simple.
During learning, people learn to label the $AB$ compound first, but they are still learning to label the $AC$ compound.
The presence of $A$ tends to push participants to generalise what they learned about $AB$, so they label $AC$ as common, which results in an error.
After making this error, attention relocates towards the uniquely predictive feature $C$ to reduce future errors.
This results in $C$ acquiring higher attentional salience than $B$.
When the ambiguous $BC$ compound is presented, this attentional allocation persists and thus $C$ will dominate responding.
This results in an irrational tendency to respond with the rare label.
According to these models, this irrationality results from an optimisation process that tries to reduce the errors people make.
This process creates an asymmetric cognitive representation that can be summarised as $AB$ belongs to common, $AB \to common$, and C belongs to rare, $C \to rare$ \cite{kruschke2001inverse}.

@@

In this work, we intend to test this basic assumption of these theories.
In the following two experiments, we will gradually remove components from the design traditionally associated with prediction error.
Our overarching goal is to investigate whether we can still observe the IBRE, even if we experimentally remove a crucial assumption of already existing accounts.
In our first attempt, building on the observational learning condition of Experiment 2 in \citeA{johansen2007paradoxical}, we implement the canonical IBRE design with a caveat that category labels are presented in unison with features.

@@

To our knowledge, there is only one attempt to implement the standard trial-by-trial IBRE procedure without explicit feedback.
In terms of a clear observational-learning version of the IBRE, \citeA{johansen2007paradoxical} included the result of a short pilot experiment in their Appendix.
Unfortunately, there is no statistical analysis confirming that the IBRE is reliably observed.
\citeA{johansen2007paradoxical} report a sample size of 16.
If we use an effect size of $d = 0.46$ observed by \citeA{inkster2022neural} and an $\alpha$ of 0.05 with a non-directional alternative hypothesis, the experiment has 24\% power\footnote{We used the method provided by the R package pwr \cite{champely2020pwr} to calculate power.}.
Given this information, the study is underpowered.
There are also no details about the procedure of this experiment.
Therefore, we cannot make direct comparisons.

@@

Additionally, there are at least three studies which directly look at error-driven processes in the IBRE.
\citeA{don2019learned}, demonstrated that on $AC$ trials, people fixated on $C$ longer than on $A$ both pre-responding during stimulus presentation and post-responding during feedback (see also \citeA{kruschke2005eye}).
This fixation bias increased with more training.
They also observe greater fixation on $C$ on $AC$ trials, relative to $B$ on $AB$ trials.
Furthermore, \citeA{wills2014attention} in an EEG study observed posterior selection negativity and concurrent frontal positivity for C relative to B, which gave evidence for an error-driven selective attentional learning process.
These studies gave evidence that attentional reallocation occurs in line with the mechanisms of EXIT-like models.
\citeA{inkster2022neural} carried out a direct investigation into brain regions underlying error-driven learning in the IBRE.
Their region of interest (ROI) analysis explicitly targeted areas that were hypothesised to be involved in the computation of prediction error.
They showed that these areas exhibited greater activation during the test phase for $C$ relative to $B$ with a presence of a shared cue during training.
Given these findings, it is reasonable to suggest that prediction-error-driven attentional reallocation occurs in a standard supervised learning paradigm and it is driven by prediction error.

@@

Table \ref{tab:abstract-exp1} summarises the abstract design of the experiment.
This design is the simplest implementation of the IBRE procedure to date.
Participants completed two phases: a training and a test phase.
In the training phase, they encountered descriptions of people, the symptoms they experienced, and their respective diseases.
These descriptions appeared in the format of '\textit{John has fever and rash, which belongs to disease Z}'.
Participants studied these examples and when they were ready to move on, they pressed the spacebar.
They needed to complete reading the description within 5 seconds.
If the 5 seconds threshold was passed, a screen appeared with the message '\textit{Please respond faster!}'.
In each training block, participants encountered 6 common diseases (common category exemplars) and 2 rare diseases (rare category exemplars).
After the second block of training, participants were given a choice.
They could either move straight to the test phase or complete another training block.
A prompt appeared saying that '\textit{Now you have the option to skip the rest of the training phase and move straight to the test phase. If you think you need some more time, you can continue training and study more patients.}'.
There were a maximum of 5 blocks they could complete.

@@

Thus the current study strongly confirms that the IBRE can be observed in an observational procedure.
In the current experimental design, the IBRE emerged in the absence of an explicit prediction error that drives the development of attentional allocation.
% This prediction error also drives the development of an asymmetric cognitive representation.
All EXIT-like theories of the IBRE rely on the assumption that this irrational rare preference arises as a result of optimising accuracy during the training phase.
In the absence of this explicit prediction error, EXIT-like theories cannot predict the presence of the IBRE.

One aspect of the current design is that participants might still experience internally-generated prediction errors from feature to categories on a trial-by-trial basis.
Given that the general assumption is that diseases cause symptoms, participants could likely assume a causal link between symptoms and diseases.
This assumed causal relationship can encourage participants to make not an explicit (responding with the category label via the keyboard) but a silent prediction.
Informally, participants might think of a certain feature--label causal relationship while reading the sentences.
People then resolve errors between the expected and the observed feature--label causality by allocating attention to rare features to distinguish diseases.

@@

\begin{table}[!ht]
  \begin{center}
    \caption{Abstract design of Experiment 2 including both test and training phases. X and Y are in place of the category labels. During the test phase, participants needed to select either X or Y to complete the features shown below.\\}
    \label{tab:abstract-exp2}
    \begin{tabular}{llr} % text alignments
      \textbf{Training (Relative Frequencies)} & \textbf{Test}& \\
      \hline
      % & \\
      ABX x 3 &  A, B, C,        & \\
      ACY x 1 &  AB, AC, BC      & x 20 \\
      \hline
    \end{tabular}
  \end{center}
\end{table}

@@

In the test phase, participants were shown \textit{incomplete} arrangement of geometric shapes and were asked to complete them.
On each test trial, they were asked to select either \textbf{X} or \textbf{Y} to complete the arrangement.
Similar to Experiment 1, each test item (incomplete arrangement of shapes) appeared 20 times.
Various arrangement of shapes appeared in the middle of the screen.
The response options X and Y with the corresponding shapes were shown below.
The prompt asked participants to pick one of the shapes to complete the arrangement.
Participants could respond by pressing either X or Y on the keyboard.
The test phase was composed of 120 trials presented across 5 blocks of 24 trials.
\\

@@

There are alternative theories of the IBRE that do not rely on processes that calculate prediction error.
A version of the dissimilarity-similarity generalised context model \citeA<DGCM, >{stewart2007dissimilarity} modified by \citeA{o2018model} has been proposed as an explanation of the IBRE.
From the perspective of DGCM, the main processes behind the rare preference are a combination of attention, memory strenght of exemplars and dissimilarity from exemplars (stored category representations).
$BC \to rare$ arises due to the combination of two factors: the high salience of $C$ relative to $B$ and the impact of the dissimilarity of $BC$ to the common category exemplar on the decision process.
Nonetheless, the model can accommodate these results only informally because it does not specify the mechanisms which encode information and produce the attentional values of each stimulus.
DGCM is a model of the test phase.
In that sense, the current experimental manipulations cannot be represented in the model.
This is the same challenge we encountered with process models - the specifications of these theories are unable to incorporate the changes to the experimental procedure.

@@

Across two experiments, we investigated whether the demonstration of the IBRE requires the "guess-and-correct" components of the standard experimental procedure.
In Experiment 1, we conducted a successful conceptual replication of \citeA{johansen2007paradoxical}, which gave evidence for the IBRE being independent of supervised learning procedures.
In addition, Experiment 2 further suggests that the IBRE generalises beyond simple predictive-learning \cite<e.g. >{medin1988problem,kruschke1996base,wills2014attention} and decision-making \cite{johansen2007paradoxical} paradigms.
This further suggests prediction error in terms of explicit feedback is not a necessary condition.
Theories of IBRE are inadequate to account for these findings, largely because of their inability to extend beyond supervised learning.

@@

We have made available the two experiments written in javascript, the analysis code, the raw data, and all other supplementary materials.
Experiment 1 is shared via \href{https://osf.io/auwvt/?view_only=2dc8384074fa4bcf9f2e3937fdaee2b4}{OSF}, and \href{TBA}{GitHub}.
Experiment 2 is shared via \href{https://osf.io/2tmc4/?view_only=489ebc888ef84a6b9b904072cfbe74df}{OSF} and \href{TBA}{Github}.
The main repository that includes this manuscript and links to the materials for the two experiments can be found on \href{TBA}{GitHub}.
Links will be inserted when document is deanonymized.

>>>>

The most prominent theories of the IBRE involve an an attentional meachanism that drives both learning and responding.
These theories are formal models.
They are: a neural network with exemplar-mediated attention to distinctive input, EXIT \cite{kruschke2001toward}, a three-layer neural network with competitive attentional gating, and a four-layer neural network with an additional rapid attentional shift \cite{paskewitz2020dissecting}.
All these explanations rely on a process that relocates attention in response to prediction errors - they update attentional values according to gradient descent.
Their explanation is simple.
During learning, people learn to label the $AB$ compound first, but they are still learning to label the $AC$ compound.
The presence of $A$ tends to push participants to generalise what they learned about $AB$, so they label $AC$ as common, which results in an error.
After making this error, attention relocates towards the uniquely predictive feature $C$ to reduce future errors.
This results in $C$ acquiring higher attentional salience than $B$.
When the ambiguous $BC$ compound is presented, this attentional allocation persists and thus $C$ will dominate responding.
This results in an irrational tendency to respond with the rare label.
According to these models, this irrationality results from an optimisation process that tries to reduce the errors people make.
This process creates an asymmetric cognitive representation that can be summarised as $AB$ belongs to common, $AB \to common$, and C belongs to rare, $C \to rare$ \cite{kruschke2001inverse}.

@@

In this work, we intend to test this basic assumption of these theories.
In the following two experiments, we will gradually remove components from the design traditionally associated with prediction error.
Our overarching goal is to investigate whether we can still observe the IBRE, even if we experimentally remove a crucial assumption of already existing accounts.
In our first attempt, we implement the canonical IBRE design with a caveat that category labels are presented in unison with features.

@@

To our knowledge, there is only one attempt to implement the standard trial-by-trial IBRE procedure without explicit feedback.
In terms of a clear observational-learning version of the IBRE, \citeA{johansen2007paradoxical} included the result of a short pilot experiment in their Appendix.
Unfortunately, there is no statistical analysis confirming that the IBRE is reliably observed.
\citeA{johansen2007paradoxical} report a sample size of 16.
If we use an effect size of $d = 0.46$ observed by \citeA{inkster2022neural} and an $\alpha$ of 0.05 with a non-directional alternative hypothesis, the experiment has 24\% power\footnote{We used the method provided by the R package pwr \cite{champely2020pwr} to calculate power.}.
Given this information, this pilot experiment is underpowered.
There are also no details about the procedure of this experiment.
Therefore, we cannot make direct comparisons.

@@

Additionally, there are at least three studies which directly look at error-driven processes in the IBRE.
\citeA{don2019learned}, demonstrated that on $AC$ trials, people fixated on $C$ longer than on $A$ both pre-responding during stimulus presentation and post-responding during feedback (see also \citeA{kruschke2005eye}).
This fixation bias increased with more training.
They also observe greater fixation on $C$ on $AC$ trials, relative to $B$ on $AB$ trials.
Furthermore, \citeA{wills2014attention} in an EEG study observed posterior selection negativity and concurrent frontal positivity for C relative to B, which gave evidence for an error-driven selective attentional learning process.
These studies gave evidence that attentional reallocation occurs in line with the mechanisms of EXIT-like models.
\citeA{inkster2022neural} carried out a direct investigation into brain regions underlying error-driven learning in the IBRE.
Their region of interest (ROI) analysis explicitly targeted areas that were hypothesised to be involved in the computation of prediction error.
They showed that these areas exhibited greater activation during the test phase for $C$ relative to $B$.
Given these findings, it is reasonable to suggest that prediction-error-driven attentional reallocation occurs in a standard supervised learning paradigm and it is driven by prediction error.

@@

% FIXME: training trials ended after 10 seconds
Table \ref{tab:abstract-exp1} summarises the abstract design of the experiment.
This design is the simplest implementation of the IBRE procedure to date.
Participants completed two phases: a training and a test phase.
In the training phase, they encountered descriptions of people, the symptoms they experienced, and their respective diseases.
These descriptions appeared in the format of '\textit{John has fever and rash, which belongs to disease Z}'.
Participants studied these examples and when they were ready to move on, they pressed the spacebar.
They needed to complete reading the description within 5 seconds.
If the 5 seconds threshold was passed, a screen appeared with the message '\textit{Please respond faster!}'.
In each training block, participants encountered 6 common diseases (common category exemplars) and 2 rare diseases (rare category exemplars).
After the second block of training, participants were given a choice.
They could either move straight to the test phase or complete another training block.
A prompt appeared saying that '\textit{Now you have the option to skip the rest of the training phase and move straight to the test phase. If you think you need some more time, you can continue training and study more patients.}'.
There were a maximum of 5 blocks they could complete.

@@

Thus the current study strongly confirms that the IBRE can be observed in an observational procedure.
In the current experimental design, the IBRE emerged in the absence of an explicit prediction error that drives the development of attentional allocation.
All EXIT-like theories of the IBRE rely on the assumption that this irrational rare preference arises as a result of optimising accuracy during the training phase.
In the absence of this explicit prediction error, EXIT-like theories cannot predict the presence of the IBRE.

One aspect of the current design is that participants might still experience internally-generated prediction errors from feature to categories on a trial-by-trial basis.
Given that the general assumption is that diseases cause symptoms, participants could likely assume a causal link between symptoms and diseases.
This assumed causal relationship can encourage participants to make not an explicit but a silent prediction.
Informally, participants might think of a certain feature--label causal relationship while reading the sentences.
People then resolve errors between the expected and the observed feature--label causality by allocating attention to rare features to distinguish diseases.

@@

\begin{table}[!ht]
  \begin{center}
    \caption{Abstract design of Experiment 2 including both test and training phases. X and Y are in place of the category labels common and rare. During the test phase, participants needed to select either X or Y to complete the features shown below.\\}
    \label{tab:abstract-exp2}
    \begin{tabular}{llr} % text alignments
      \textbf{Training (Relative Frequencies)} & \textbf{Test}& \\
      \hline
      % & \\
      ABX x 3 &  A, B, C,        & \\
      ACY x 1 &  AB, AC, BC      & x 20 \\
      \hline
    \end{tabular}
  \end{center}
\end{table}

@@

In the test phase, participants were shown \textit{incomplete} arrangements of geometric shapes and were asked to complete them.
On each test trial, they were asked to select either \textbf{X} or \textbf{Y} to complete the arrangement.
Similar to Experiment 1, each test item (incomplete arrangement of shapes) appeared 20 times.
Various arrangement of shapes appeared in the middle of the screen.
The response options X and Y with the corresponding shapes were shown below.
The prompt asked participants to pick one of the shapes to complete the arrangement.
Participants could respond by pressing either X or Y on the keyboard.
The test phase was composed of 120 trials presented across 5 blocks of 24 trials.
\\

@@

There are alternative theories of the IBRE that do not rely on processes that calculate prediction error.
A version of the dissimilarity-similarity generalised context model \citeA<DGCM, >{stewart2007dissimilarity} modified by \citeA{o2018model} has been proposed as an explanation of the IBRE.
From the perspective of DGCM, the main processes behind the rare preference are a combination of attention, memory strength of exemplars and dissimilarity from exemplars (stored category representations).
$BC \to rare$ arises due to the combination of two factors: the high salience of $C$ relative to $B$ and the impact of the dissimilarity of $BC$ to the common category exemplar on the decision process.
Nonetheless, the model can accommodate these results only informally because it does not specify the mechanisms which encode information and produce the attentional values of each stimulus.
DGCM is a model of the test phase.
In that sense, the current experimental manipulations cannot be represented in the model.
This is the same challenge we encountered with process models - the specifications of these theories are unable to incorporate the changes to the experimental procedure.

@@

% TODO: introduce guess-and-correct before the conclusion
Across two experiments, we investigated whether the demonstration of the IBRE requires the "guess-and-correct" components of the standard experimental procedure.
In Experiment 1, we conducted a successful conceptual replication of \citeA{johansen2007paradoxical}, which gave evidence for the IBRE being independent of supervised learning procedures.
In addition, Experiment 2 further suggests that the IBRE generalises beyond simple predictive-learning \cite<e.g. >{medin1988problem,kruschke1996base,wills2014attention} and decision-making \cite{johansen2007paradoxical} paradigms.
This further suggests prediction error in terms of explicit feedback is not a necessary condition.
Theories of IBRE are inadequate to account for these findings, largely because of their inability to extend beyond supervised learning.

@@

We have made available the two experiments written in javascript, the analysis code, the raw data, and all other supplementary materials.
Experiment 1 is shared via \href{https://osf.io/auwvt/?view_only=2dc8384074fa4bcf9f2e3937fdaee2b4}{OSF}, and \href{TBA}{GitHub}.
Experiment 2 is shared via \href{https://osf.io/2tmc4/?view_only=489ebc888ef84a6b9b904072cfbe74df}{OSF} and \href{TBA}{Github}.
The main repository that includes this manuscript and links to the materials for the two experiments can be found on \href{TBA}{GitHub}.
GitHub links will be inserted when document is deanonymized.
